Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-06-24 demystify.tex
updates

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF2.0 program}, which has helped many students in gaining admission to top CS PhD programs. VEF2.0 invites US faculty members from leading institutions to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily just the interview), these reference letters are generally effective and have helped many students get into top PhD programs in the US.

@@

\textbf{Advice}: If you haven't taken the GRE or have bad scores, then don't waste time (re)taking it. It does not help much!

@@

\textbf{Advice}: Just do well enough to pass the minimum requirement set by the university. %, which is required for TA (\autoref{sec:funding}). 
Just as with grades (\autoref{sec:your-school}) and GRE, having high scores in these tests might not help, but having too low scores can be a \red{red flag} and sometimes results in an automatic rejection (\autoref{sec:why-rejected}), e.g., below the minimum requirement.

@@

\textbf{Advice}: do not spend much time talking about non-research-related job in your SOP.
However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab, e.g., Microsoft, DoE, VinAI in Vietnam, etc (\autoref{sec:research-experiences}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potential in their LOR (\autoref{sec:lor}).

@@

\textbf{Advice}: Prepare your CV in such a way that it allows the reviewers to quickly scan to identify major achievements, e.g., Publications, Programming Competition Awards, and Teaching Experience.

@@

\textbf{Advice}: Treat the interview as an informal chat. Have elevator pitches about your research experiences and interests. You might also want to have a 5-minute presentation about your research. If a prof. asks you to read a paper, do it and be prepared to discuss it. You can also ask if you need to prepare for a coding problem. Finally, prepare some thoughtful questions to ask, e.g., about the program, the university, or the professor's research.

@@

By around March you should hear back from most PhD programs that want to admit you.  In many cases you will not hear back from schools that do not want to admit you.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will likely have to make your decision (\autoref{sec:accept-postpone-decline}) by around \emph{April 15}, which is the deadline for most schools
(\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in the DC area is likely higher than in Lincoln, Nebraska due to higher living costs). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so the stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer (\autoref{sec:summer-funding}) through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most sources of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipends, but they could have extra activity fees or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

@@

\paragraph{How to get TA?}  Unless you have other funding such as RA or fellowships, TA is typically the default for CS PhD programs. When you apply to be a full-time student,  state that you need financial assistant. Typically, adcom will either admit you and give you GTA, or reject you; i.e., we do not admit a student without supporting them.

>>>>

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF2.0 program}, which has helped many students gain admission to top CS PhD programs in the US. VEF2.0 invites US faculty members from leading institutions to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily just the interview), these reference letters are generally effective and have helped many students get into good PhD programs.

@@

Thus, if you haven't taken the GRE or have bad scores, then don't waste time (re)taking it. It does not help much!

@@

You should just do well enough to pass the minimum requirement set by the university. %, which is required for TA (\autoref{sec:funding}). 
Just as with grades (\autoref{sec:your-school}) and GRE, having high scores in these tests might not help, but having too low scores can be a \red{red flag} and sometimes results in an automatic rejection (\autoref{sec:why-rejected}), e.g., below the minimum requirement.

@@

Do not spend much time talking about non-research-related job in your SOP.
However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab, e.g., Microsoft, DoE, VinAI in Vietnam, etc (\autoref{sec:research-experiences}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potential in their LOR (\autoref{sec:lor}).

@@

Prepare your CV in such a way that it allows the reviewers to quickly scan to identify major achievements, e.g., Publications, Programming Competition Awards, and Teaching Experience.

@@

You should treat the interview as an informal chat. Have ``elevator pitches'' about your research experiences and interests. You might also want to have a 5-minute presentation about your research. If a prof. asks you to read a paper, do it and be prepared to discuss it. You can also ask if you need to prepare for a coding problem. Finally, prepare some thoughtful questions to ask, e.g., about the program, the university, or the professor's research.

@@

By around March you should hear back from most PhD programs that want to admit you.  In many cases you will not hear back from schools that do not want to admit you (see \autoref{sec:late-rejection}).
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will likely have to make your decision (\autoref{sec:accept-postpone-decline}) by around \emph{April 15}, which is the deadline for most schools
(\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in the DC area is likely higher than in Lincoln, Nebraska due to higher living costs). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so the stipend is for 9 months. Third, you might get paid over the summer (\autoref{sec:summer-funding}) through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most sources of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipends, but they could have extra activity fees or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

@@

\paragraph{How to get TA?}  Unless you have other funding such as RA or fellowships, TA is typically the default for CS PhD programs. In your application, indicate that you need financial assistant. Typically, adcom will either admit you and give you TA, or reject you. We do not admit a student without supporting them.
